### 2015-02-02 doc/thesis/evaluation.tex
Minor improvements to the text about case studies

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
\textbf{Case study} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
Lua Standard Library & 80 & 6 & 8 & 35 & 129 & 62\% \\
\hline
MD5 & 13 & 0 & 0 & 0 & 13 & 100\% \\
\hline
LuaSocket & 109 & 1 & 2 & 11 & 123 & 89\% \\
\hline
HTTP Digest & 0 & 0 & 1 & 0 & 1 & 0\% \\
\hline
Typical & 1 & 0 & 0 & 0 & 1 & 100\% \\
\hline
Modulo 11 & 7 & 0 & 0 & 2 & 9 & 78\% \\
\hline
Typed Lua Compiler & 144 & 0 & 1 & 9 & 154 & 94\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for each case study}
\label{tab:evalbycase}
\end{table}

@@

All of the modules in the standard library are implemented in C, so
we used Typed Lua to type just the interface of each module.
The \texttt{debug} module is the only one that we did not include in our
evaluation results, because it provides several functions that violate
basic assumptions about Lua code \citep{luamanual}.
For instance, we can use its function \texttt{setlocal} to change the value
of a local variable that is not visible in the current scope.
Table \ref{tab:evallsl} summarizes the evaluation results for the Lua Standard Library
(version 5.3).

@@

We could give precise static types to several members of
the Lua Standard Library, though some of them are still
hard to type with the current type system.
Now, we discuss why some modules are difficult to type,
and what kind of limitations they represent on our type system.

The \texttt{base} module was quite hard to type because it
includes several functions that have a dynamic behavior.
For instance, the function \texttt{next} traverses all the
fields of a table, but the order that the indices are
enumerated is not specified.
For this reason, it is also difficult to type the function
\texttt{pairs}, as it uses the function \texttt{next} to
traverse a table.
The functions \texttt{getmetatable}, \texttt{rawget}, and
\texttt{rawset} are other examples of functions that their
behavior depend on the C implementation.

We could not give a precise static type to \texttt{ipairs}
because our type system does not have parametric polymorphism.
This function returns an iterator function that traverses a
list of elements.

The \texttt{base} module also includes two special cases:
the functions \texttt{tonumber} and \texttt{collectgarbage}.

The function \texttt{tonumber} is a special case because the
type of the first parameter depends on the type of the
second parameter.
For instance, we can call \texttt{tonumber(1)}, but we cannot
call \texttt{tonumber(1,2)}.
More precisely, the first argument of \texttt{tonumber} can be
a value of any type if it is the only argument, but it must
be a string if there is a second argument,
which must be an integer.

The function \texttt{collectgarbage} is a special case because
its return type depends on the value of the first parameter,
which should be a string.
For instance, calling \texttt{collectgarbage("collect")} returns an integer number,
calling \texttt{collectgarbage("count")} returns a floating point number,
and calling \texttt{collectgarbage("isrunning")} returns a boolean.

@@

We could type most of the members of the \texttt{package} module,
but we could not type the following tables: \texttt{package.loaded},
\texttt{package.preload}, and \texttt{package.loaders}.
The first stores loaded modules, while the others store module \emph{loaders}.
They are difficult to type because their types are dynamic and
depend on the modules a program loads.

@@

The \texttt{string} module is not that hard to type, but it includes
some functions that require dependent types.
For instance, the type of the arguments that we pass to
\texttt{string.format} depend on the format string we are using.
It is fine to call \texttt{string.format("\%d", 1)}, but
\texttt{string.format("\%d", true)} generates a run-time error.

The \texttt{utf8} module was quite straightforward to type,
as its members are only operations over strings.

The \texttt{table} module is specially difficult to type because
almost all of its members require parametric polymorphism.
All the functions of this module either receive or return a list
of elements, and parametric polymorphism would help us to describe
them with a generic type.

@@

Even though the \texttt{math} module looks straightforward to type,
it includes several functions that the return type depend on the input type.
For instance, \texttt{math.abs(1)} returns the integer \texttt{1},
while \texttt{math.abs(1.1)} returns the floating point \texttt{1.1}.

The \texttt{math} module also includes a special case that is hard to type:
the function \texttt{math.random}.
This function is difficult to type because its type depends on the calling arity.
Calling \texttt{math.random()} returns a random floating point number
between \texttt{0} and \texttt{1}.
Calling \texttt{math.random(10)} is equivalent to \texttt{math.random(1,10)},
and returns an integer between this interval.
This function also does not follow the semantics of Lua on discarding extra arguments,
and generates a run-time error whenever we pass more than two arguments.
There is also a problem on its documentation, as it suggests that
the two integers are optional parameters, but, for \texttt{math.random},
optional parameters behave in a different way.
Usually, Lua functions replace optional parameters with a default value
when they are \texttt{nil}, but \texttt{math.random} generates a run-time
error instead.

@@

The functions \texttt{io.read} and \texttt{io.lines} are difficult
to type because their return types depend on their input types.
For instance, \texttt{io.read("l")} returns a string or \texttt{nil},
\texttt{io.read("n")} returns a number or \texttt{nil}, and calling
\texttt{io.read("l", "n")} returns a string or \texttt{nil} and a number or \texttt{nil}.
The function \texttt{io.lines}, and the methods \texttt{file:read} and
\texttt{file:lines} are difficult to type for the same reason.

The function \texttt{os.execute} is the only member of the \texttt{os}
module that we could not give a precise type.
This function is difficult to type because its return type depend on
the input type.
More precisely, \texttt{os.execute()} returns a boolean, while
\texttt{os.execute(c)} returns \texttt{(boolean?, string, number)},
where \texttt{c} is a string that represents the command to be executed.

@@

Even tough it was quite straightforward to type the MD5 library,
we found a little difference between its documentation and its behavior.
The documentation suggests that the type of the function \texttt{md5.update}
is \texttt{(md5\string_context, string) -> (md5\string_context)},
though there is a call to this function in the test script that passes
an extra string argument.
Reading the source code, we found that its actual type is
\texttt{(md5\string_context, string*) -> (md5\string_context)},
that is, we can pass zero or more strings to \texttt{md5.update}.

This case study shows that type annotations help programmers maintain the
documentation updated, as the type checker always validate them.

@@

LuaSocket is a library that adds network support to Lua,
and it is split into two parts: a core that is written in C and a set of
Lua modules.
The C core provides TCP and UDP support, while the Lua modules provide
support for SMTP, HTTP, and FTP client protocols, MIME encoding,
URL manipulation, and LTN12 filters \citep{nehab2008ltn012}.
We used Typed Lua's description files to type both parts.
This is the only case study that we test description files to statically type
a Lua module instead of rewriting it.
Table \ref{tab:evalsocket} summarizes the evaluation results for LuaSocket.

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
socket & 50 & 0 & 0 & 10 & 60 & 83\% \\
\hline
ftp & 5 & 0 & 1 & 0 & 6 & 83\% \\
\hline
http & 4 & 0 & 1 & 0 & 5 & 80\% \\
\hline
smtp & 6 & 0 & 0 & 1 & 7 & 86\% \\
\hline
mime & 17 & 0 & 0 & 0 & 17 & 100\% \\
\hline
ltn12 & 19 & 1 & 0 & 0 & 20 & 95\% \\
\hline
url & 8 & 0 & 0 & 0 & 8 & 100\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for LuaSocket}
\label{tab:evalsocket}
\end{table}

We could type most of the members in the \texttt{socket} module,
which is the C core.
However, this module includes some functions such as
\texttt{socket.try}, \texttt{socket.protect}, and \texttt{socket.skip}
that have a dynamic behavior.
This means that the return type of these functions depend on the
type of the values that we use to call them.
For instance, \texttt{socket.skip(0, nil, "hello")} returns the
tuple \texttt{(nil, "hello")}, \texttt{socket.skip(1, nil, "hello")}
returns only the string \texttt{"hello"}, and
\texttt{socket.skip(-1, nil, "hello")} returns the tuple
\texttt{(-1, nil, "hello")}.
Passing a negative number to \texttt{socket.skip} can be
dangerous, as it returns anything that might be in the stack.
For instance, \texttt{f = socket.skip(-2)} assigns \texttt{socket.skip}
to \texttt{f}.
Our type system cannot handle the type of negative numbers, as
this requires the refinement types from hybrid type checking \citep{flanagan2006htc}.

We could type most of the members of the modules \texttt{ftp},
\texttt{http}, and \texttt{smtp}, but we could not precisely type
the functions \texttt{ftp.get}, \texttt{http.request}, and
\texttt{smtp.message}.

The function \texttt{ftp.get} downloads data from a given URL,
and it is difficult to type because its return type depends on
the type of its input parameter.
More precisely, \texttt{ftp.get(url)} returns the tuple
\texttt{(string) | (nil, string)} if \texttt{url} is a string,
and it returns the tuple \texttt{(number) | (nil, string)} if
\texttt{url} is a table.

The function \texttt{http.request} downloads data from a given URL,
and it is difficult to type because its return type depends on
the types of its input parameters.
More precisely, \texttt{http.request(url, body)} returns the tuple
\texttt{(string, number, \{string:string\}, number) | (nil, string)}
if \texttt{url} is a string and \texttt{body} is another string or \texttt{nil},
but it returns the tuple
\texttt{(number, number, \{string:string\}, number) | (nil, string)}
if \texttt{url} is a table and \texttt{body} is \texttt{nil}.

The function \texttt{smtp.message} creates a function that sends
an SMTP message body, and it is difficult to type because of the
type of the message body, which is recursive and makes harder
its initialization.

The modules \texttt{mime} and \texttt{ltn12} have a strong connection.
The \texttt{mime} module offers low-level and high-level filters
that apply and remove some text encodings.
The low-level filters are written in C, while the high-level filters
use the function \texttt{ltn12.filter.cycle} along with the low level
filters to create standard filters.

Even though we could type all the members of the \texttt{mime} module,
the function \texttt{ltn12.filter.cycle} is the only member of the
\texttt{ltn12} module that we could not give a precise type.
This function is difficult to type because the \texttt{mime} low-level
filters are polymorphic.

The \texttt{url} module provides functions that manipulate URLs.
It is straightforward to type, although it defines a table type for
parsed URLs that is not precise enough.
This type is problematic because it is a record type where all
fields are optional, and makes any table type that does not include
the optional fields statically type check.

This case study shows that abusing of dynamic typing usually impacts
the users of a module, as they have to keep on checking each return
value before using it to prevent API misuses.

@@

It is difficult to type the interface of the \texttt{http-digest} module
because it is an extension to the \texttt{http} module from LuaSocket.
The \texttt{http-digest} module only exports the function
\texttt{http-digest.request}, which extends the function
\texttt{http.request} with MD5 authentication.
Like \texttt{http.request}, the return type of \texttt{http-digest.request}
also depends on the types of its input parameters.

Even though we could not precisely type the interface that \texttt{http-digest}
exports, we could use only static types to rewrite it, and they pointed a bug
in the code.
The problem was related to the way the library was loading the MD5 library
that should be used. 
This part of the code checks the existence of three different MD5 libraries
in the system, and uses the first one that is available, or generates an
error when none is available.
The code that loads the first option was fine, but the code that loads the
second and third options were trying to access an undefined global variable.

@@

However, we hit some limitations of our type system to rewrite this module.
First, it uses the function \texttt{getmetatable} to get a table and
checks whether this table has the field \texttt{\string_\string_type} or not.
We could not give a precise type to \texttt{getmetatable}, so we used the dynamic
type \texttt{any} as its return type, and this generates a warning.
Second, the module returns a metatable that extends \texttt{\string_\string_call}
with \texttt{typical.type}, that is, we can use the module itself as a function,
though it is a table.
Our type system still does not support metatables, so we did not extend our version
of the \texttt{typical} module to support \texttt{\string_\string_call}.
Third, the module uses \texttt{ipairs} to iterate over an array of functions,
but our type system also has limited support to \texttt{ipairs}, and generates
a warning when we try to use the indexed value inside the \texttt{for} body.
We removed this warning using the numeric \texttt{for} to perform the same loop.

@@

Even though we had these two issues to rewrite the \texttt{mod11} module,
we could use only static types to rewrite it, and found some interesting
points.
The code relies a lot on implicit conversions between strings and numbers,
and some parts of the code keep on changing the type of local variables.
These are two practices that may hide bugs.

@@

The \texttt{tllexer} module defines common lexical rules for
the Typed Lua parser and the description file parser.
This module is hard to type because it exports LPeg \citep{lpeg,ierusalimschy2009lpeg}
patterns.
For instance,
\begin{verbatim}
    tllexer.Number = lpeg.C(Hex + Float + Int) / tonumber
\end{verbatim}
exports the pattern that matches a number.
Note that LPeg overloads the definition of arithmetic operators,
a feature that our type system does not support yet.
That is the reason why we could not precisely type the LPeg
patterns that this module exports.
However, we could type the two error reporting functions that this
module exports.

@@

Traversing the AST would also be problematic if we had not included
a way to decompose unions of table types, as we mentioned in Section \ref{sec:alias}.

Bootstraping the compiler also helped revealing some bugs.
We found some accesses to undeclared global variables and also
to undeclared table fields.
The compiler also helped pointing the places where we should narrow
a nilable value before using it.
In fact, this point appeared in all the case studies that we rewrote
the Lua code.
This means that Lua programmers often use values before checking
whether it is \texttt{nil}.

>>>>

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
\textbf{Case study} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
Lua Standard Library & 80 & 6 & 8 & 35 & 129 & 62\% \\
\hline
MD5 & 13 & 0 & 0 & 0 & 13 & 100\% \\
\hline
LuaSocket & 110 & 1 & 2 & 10 & 123 & 89\% \\
\hline
HTTP Digest & 0 & 0 & 1 & 0 & 1 & 0\% \\
\hline
Typical & 1 & 0 & 0 & 0 & 1 & 100\% \\
\hline
Modulo 11 & 7 & 0 & 0 & 2 & 9 & 78\% \\
\hline
Typed Lua Compiler & 144 & 0 & 1 & 9 & 154 & 94\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for each case study}
\label{tab:evalbycase}
\end{table}

@@

All of the modules in the standard library are implemented in C, so
we used Typed Lua to type just the interface of each module.
The \texttt{debug} module is the only one that we did not include in our
evaluation results, because it provides several functions that violate
basic assumptions about Lua code \citep{luamanual}.
For instance, we can use the function \texttt{debug.setlocal} to change the value
of a local variable that is not visible in the current scope.
Table \ref{tab:evallsl} summarizes the evaluation results for the Lua Standard Library
(version 5.3).

@@

We could give precise static types to several members of
the Lua Standard Library, though some of them are still
hard to type with our current type system.
Now we will discuss why some modules are difficult to type,
and what kind of limitations they represent in our type system.

The \texttt{base} module was quite hard to type because it
includes several functions that rely on reflection.
For instance, the function \texttt{pairs} traverses all
keys and values that are stored in a given table.
As another example, the function \texttt{getmetatable}
returns the metatable of a given table.

We could not give a precise static type to \texttt{ipairs},
because our type system does not have parametric polymorphism.
This function returns an iterator function that traverses a
list of elements.

The \texttt{base} module also includes two overloaded functions:
\texttt{tonumber} and \texttt{collectgarbage}.

In the case of \texttt{tonumber}, the type of the first parameter
depends on the type of the second parameter.
For instance, we can call \texttt{tonumber(1)}, but we cannot
call \texttt{tonumber(1,2)}.
More precisely, the first argument of \texttt{tonumber} can be
a value of any type if it is the only argument, but it must
be a string if there is a second argument,
which must be an integer.

In the case of \texttt{collectgarbage}, the return type changes
according to an input literal string.
For instance, calling \texttt{collectgarbage("collect")} returns an integer,
calling \texttt{collectgarbage("count")} returns a floating point,
and calling \texttt{collectgarbage("isrunning")} returns a boolean.

@@

We could type most of the members of the \texttt{package} module,
but we could not type the following tables: \texttt{package.loaded},
\texttt{package.preload}, and \texttt{package.loaders}.
The first stores loaded modules, while the others store module \emph{loaders}.
They are difficult to type because their types rely on reflection,
that is, their types depend on the modules a program loads.

The \texttt{string} module was not that hard to type, but it includes
some functions that we could not type, as they rely on format
strings.
For instance, the type of the arguments that we pass to
\texttt{string.format} depend on the format string we are using.
It is fine to call \texttt{string.format("\%d", 1)}, but
\texttt{string.format("\%d", true)} raises a run-time error.

@@

The \texttt{utf8} module was straightforward to type,
as its members are only operations over strings.

The \texttt{table} module was specially difficult to type because
most of its functions require parametric polymorphism.
These functions either receive or return a list of elements, and
parametric polymorphism would help us to describe them with a generic type.

@@

Even though the \texttt{math} module looks straightforward to type,
it includes several overloaded functions.
For instance, \texttt{math.abs(1)} returns the integer \texttt{1},
while \texttt{math.abs(1.1)} returns the floating point \texttt{1.1}.

The \texttt{math} module also includes a function that is hard to type:
the function \texttt{math.random}.
It is difficult to type because its type depends on the calling arity.
Calling \texttt{math.random()} returns a random floating point
between \texttt{0} and \texttt{1}.
Calling \texttt{math.random(10)} is equivalent to \texttt{math.random(1,10)},
and returns an integer between this interval.
This function also does not follow the semantics of Lua on discarding extra arguments,
and generates a run-time error whenever we pass more than two arguments.

@@

The functions \texttt{io.read} and \texttt{io.lines} are difficult
to type because they are overloaded functions.
For instance, \texttt{io.read("l")} returns a string or \texttt{nil},
\texttt{io.read("n")} returns a number or \texttt{nil}, and calling
\texttt{io.read("l", "n")} returns a string or \texttt{nil} and a number or \texttt{nil}.
The function \texttt{io.lines}, and the methods \texttt{file:read} and
\texttt{file:lines} are difficult to type for the same reason.

The function \texttt{os.execute} is the only member of the \texttt{os}
module that we could not give a precise type, as it is an overloaded
function.
More precisely, \texttt{os.execute()} returns a boolean, while
\texttt{os.execute(c)} returns \texttt{(boolean?, string, number)},
where \texttt{c} is a string that represents the command to be executed.

@@

Even tough it was straightforward to type the MD5 library,
we found a little difference between its documentation and its behavior.
The documentation suggests that the type of the function \texttt{md5.update}
is \texttt{(md5\string_context, string) -> (md5\string_context)},
though there is a call to this function in the test script that passes
an extra string argument.
Reading the source code, we found that its actual type is
\texttt{(md5\string_context, string*) -> (md5\string_context)},
that is, we can pass zero or more strings to \texttt{md5.update}.

This case study shows that type annotations help programmers maintain the
documentation updated, as the type checker always validates them.

@@

LuaSocket is a library that adds network support to Lua,
and it is split into two parts: a core that is written in C and a set of
Lua modules.
The C core provides TCP and UDP support, while the Lua modules provide
support for SMTP, HTTP, and FTP client protocols, MIME encoding,
URL manipulation, and LTN12 filters \citep{nehab2008ltn012}.
We used Typed Lua's description files to type both parts, as we also
wanted to use LuaSocket to test description files to statically type
the interface of modules that are written in Lua.
Table \ref{tab:evalsocket} summarizes the evaluation results for LuaSocket.

\begin{table}[!ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Module} & \textbf{easy} & \textbf{poly} & \textbf{inter} & \textbf{hard} & \textbf{Total} & \textbf{\%} \\
\hline
socket & 50 & 0 & 0 & 10 & 60 & 83\% \\
\hline
ftp & 5 & 0 & 1 & 0 & 6 & 83\% \\
\hline
http & 4 & 0 & 1 & 0 & 5 & 80\% \\
\hline
smtp & 7 & 0 & 0 & 0 & 7 & 100\% \\
\hline
mime & 17 & 0 & 0 & 0 & 17 & 100\% \\
\hline
ltn12 & 19 & 1 & 0 & 0 & 20 & 95\% \\
\hline
url & 8 & 0 & 0 & 0 & 8 & 100\% \\
\hline
\end{tabular}
\end{center}
\caption{Evaluation results for LuaSocket}
\label{tab:evalsocket}
\end{table}

We could type most of the members in the \texttt{socket} module,
which is the C core.
However, this module includes some functions such as
\texttt{socket.try}, \texttt{socket.protect}, and \texttt{socket.skip}
that are hard to type because they rely on reflection.
For instance, we can use \texttt{socket.skip} to choose the number of
values that we want to return.
As an example, calling \texttt{socket.skip(1, nil, "hello")}
returns only the string \texttt{"hello"}, because \texttt{1} indicates
that we do not want to return the first value.
Passing a negative number to \texttt{socket.skip} can be
dangerous, as it returns anything that might be in the stack.
As an example, calling \texttt{socket.skip(-1, nil, "hello")}
returns the tuple \texttt{(-1, nil, "hello")}, because \texttt{-1} makes
\texttt{socket.skip} does not skip any values.
As another example, \texttt{f = socket.skip(-2)} assigns \texttt{socket.skip}
to \texttt{f}, as \texttt{-2} gets \texttt{socket.skip} from the stack.
Our type system cannot handle the type of negative numbers, as
this requires more complex types such as the refinement types from
hybrid type checking \citep{flanagan2006htc}.

We could type most of the members of the modules \texttt{ftp} and
\texttt{http}, but we could not precisely type
the functions \texttt{ftp.get} and \texttt{http.request} because
they are overloaded functions.

The function \texttt{ftp.get} downloads data from a given URL,
which can be either a string or a table.
More precisely, \texttt{ftp.get(url)} returns the tuple
\texttt{(string) | (nil, string)} if \texttt{url} is a string,
and it returns the tuple \texttt{(number) | (nil, string)} if
\texttt{url} is a table.

The function \texttt{http.request} downloads data from a given URL,
which can be either a string or a table.
More precisely, \texttt{http.request(url, body)} returns the tuple
\texttt{(string, number, \{string:string\}, number) | (nil, string)}
if \texttt{url} is a string and \texttt{body} is another string or \texttt{nil},
but it returns the tuple
\texttt{(number, number, \{string:string\}, number) | (nil, string)}
if \texttt{url} is a table and \texttt{body} is \texttt{nil}.

The modules \texttt{mime} and \texttt{ltn12} have a strong connection.
The \texttt{mime} module offers low-level and high-level filters
that apply and remove some text encoding.
The low-level filters are written in C, while the high-level filters
use the function \texttt{ltn12.filter.cycle} along with the low level
filters to create standard filters.

Even though we could type all the members of the \texttt{mime} module,
the function \texttt{ltn12.filter.cycle} is the only member of the
\texttt{ltn12} module that we could not give a precise type.
This function is difficult to type because it is polymorphic.

The modules \texttt{smtp} and \texttt{url} were straightforward to type.
The \texttt{smtp} module provides functions that send e-mails.
The \texttt{url} module provides functions that manipulate URLs.

@@

It is difficult to type the interface of the \texttt{http-digest} module
because it is an extension to the \texttt{http} module from LuaSocket.
The \texttt{http-digest} module only exports the function
\texttt{http-digest.request}, which extends the function
\texttt{http.request} with MD5 authentication.
Like \texttt{http.request}, \texttt{http-digest.request}
is also an overloaded function.

Even though we could not precisely type the interface that \texttt{http-digest}
exports, we could use only static types to annotate it, and they pointed a bug
in the code.
The problem was related to the way the library was loading the MD5 library
that should be used. 
This part of the code checks the existence of three different MD5 libraries
in the system, and uses the first one that is available, or generates an
error when none is available.
The code that loads the first option was fine, but the code that loads the
second and third options were trying to access an undefined global variable.

@@

However, we hit some limitations of our type system while annotating this module.
First, it uses the function \texttt{getmetatable} to get a table and
checks whether this table has the field \texttt{\string_\string_type} or not.
We could not give a precise type to \texttt{getmetatable}, so we used the dynamic
type \texttt{any} as its return type, and this generates a warning.
Second, the module returns a metatable that extends \texttt{\string_\string_call}
with \texttt{typical.type}, that is, we can use the module itself as a function,
though it is a table.
Our type system still does not support metatables, so we did not extend our version
of the \texttt{typical} module to support \texttt{\string_\string_call}.
Third, the module uses \texttt{ipairs} to iterate over an array of functions,
but our type system also has limited support to \texttt{ipairs}, and generates
a warning when we try to use the indexed value inside the \texttt{for} body.
We removed this warning using the numeric \texttt{for} to perform the same loop.

@@

Even though we had these two issues to annotate the \texttt{mod11} module,
we could use only static types to annotate it, and we found some interesting
points.
The code relies on implicit conversions between strings and numbers,
and some parts of the code keep on changing the type of local variables.
These are two practices that may hide bugs.

@@

The \texttt{tllexer} module defines common lexical rules for
the Typed Lua parser and the description file parser.
This module is hard to type because it uses LPeg \citep{lpeg,ierusalimschy2009lpeg}
patterns, and LPeg uses overloaded arithmetic operators to
build LPeg patterns. 
We could not precisely type LPeg patterns because our
type system still does not support overloading arithmetic
operators.
In this module, we could only type the two error reporting functions
that it exports.

@@

Traversing the AST would also be problematic if we had not included
a way to discriminate unions of table types, as we mentioned in Section \ref{sec:alias}.

Bootstraping the compiler also helped revealing some bugs.
We found some accesses to undeclared global variables and also
to undeclared table fields.
The compiler also helped pointing the places where we should narrow
a nilable value before using it.
In fact, this point appeared in all the case studies that we used
Typed Lua to annotate Lua code.
This means that Lua programmers often use possibly \texttt{nil} values
before checking whether it is \texttt{nil}.
